---
title: "Credit Modelling"
author: "Chris Hua"
date: "11/7/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(magrittr)
library(broom)
library(ggplot2)
insample <- read.csv("File1_IS_data.csv")
insample %>% select(default) %>% table
```

# Notes

Chris Hua, Kevin Huo, Arjun Jain, Juan Manubens

# Basic Modelling

We model "default probability" with respect to "FICO" score only. First, we note the distribution of the outcome variable:

```{r}
insample %>% select(default) %>% table
```

As well as the distribution of the FICO score, among applicants in this dataset:

```{r}
insample %>%
    ggplot(aes(fico)) + geom_histogram() + 
    theme_classic() + 
    ggtitle("")
```


**a.** We expect a negative coefficient on FICO. Intuitively, a customer is less likely to default the better their credit score is. We would also expect a positive intercept, because for a customer with 0 credit score (the condition for the intercept), they have terrible chances of paying off the bill, and thus high chance for default

**b.** We estimate our model using `glm` approach with a binomial prior.

```{r}
insample %<>%
    mutate(default = default == "Defaulted")

log_fit <- glm(default ~ fico, data = insample, family = "binomial")
log_fit %>% tidy %>% knitr::kable(digits = 2)
```

By p-value, `fico` is a statistically significant value at better than at 0.01 level, indicating that by itself, `fico` is a useful predictor for default outcomes.

The intercept is strongly positive, which indicates that default is very likely for somebody with a `fico` score of 0. Alternatively, the intercept suggests that somebody with a score of 768 has 0 chance of default. The positive coefficient means that for each point of FICO score, the likelihood of a negative outcome decreases by 0.01. Each of these observations lines up with our intuitions.

# Model Evaluation

```{r}
predict(log_fit, insample) %>% sort %>% plot
```

